Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-08-16 extended_catalog/paper/extended_catalog.tex
before (most of) keiths changes

@@

@@

% some units
\newcommand{\kev}{\text{Kev}\xspace}
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\tev}{\text{TeV}\xspace}
\newcommand{\sr}{\text{sr}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\renewcommand{\sec}{\text{s}\xspace}
\newcommand{\tsext}{{\ensuremath{\text{TS}_{\text{ext}}}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_{\text{inc}}}\xspace}
\newcommand{\loglikelihood}{\ensuremath{LL}\xspace}
\newcommand{\chandra}{\text{{\em Chandra}}\xspace}
\newcommand{\rosat}{\text{{\em ROSAT}}\xspace}
\newcommand{\asca}{\text{{\em ASCA}}\xspace}
\newcommand{\einstein}{\text{{\em Einstein}}\xspace}
\newcommand{\xmmnewton}{\text{{\em XMM-Newton}}\xspace}

@@

Using two years of all sky surveying data, the LAT Collaboration published
a catalog of 1873 \gev sources\cite{second_cat}. Many of these can be
positionally associated with a variety of source classes including Active
Galactic Nuclei (AGN), pulsars, galaxies, and supernova remnants (SNRs).
While many of the LAT sources can be spatially resolved when observed
at other frequencies, detection of an extension at \gev energies by
the LAT is complicated by the rather large point-spread function (PSF)
of the instrument. The angular resolution for events converting in the
front part of the detector, defined as the 68\% containment radius for
a point source, is approximately $3.5^{\circ}$ at 100 \mev, improving
to about 0.1$^{\circ}$ at 10 \gev.

However, a number of sources positionally coincident with LAT sources
exhibit extension at other wavelengths which are larger or comparable to
the LAT PSF.  In particular certain SNRs, pulsar wind nebulae (PWNe),
molecular clouds, normal galaxies, galaxy clusters or dark matter
satellites can therefore be expected to be observed as spatially extended
sources in the LAT.  The LAT collaboration has previously reported on the
discovered of five spatially extended supernova remnants, IC443, W28, W44,
W51C, and RX J1713.7-3946 (\cite{ic443,w28,w44,w51c,rx_j1713_lat}). In
addition, three extended PWNe were detected: MSH 15-52, Vela~X,
and HESS\,J1825-137 (\cite{msh1552,velax,fermi_hess_j1825}), as well
as two close-by galaxies, the Large Magellanic and Small Magalenic
Cloud (\cite{lmc,smc}), and finally the radio galaxy Centarus A
(\cite{cen_a_lat}) .

For each of these sources, the extension was determined based on
dedicated analysis of the particular source or region of interest
instead of being the outcome of a systematic search of the LAT catalog.
For the systematic scan of all LAT catalog sources presented here, we
therefore expect to find additional spatially extended LAT sources, in
particular since many extended sources in the Galactic plane have been
detected at \tev energies using Air Cherenkov detectors and \tev and
\gev emission often originates from the same population of high-energy
particles.  A variety of spatially extended sources have been detected
in the Galactic plane at \tev energies. Most prominent was a survey of
the Galactic plane using the High Energy Stereoscopic System (H.E.S.S)
which presented 14 spatially extended sources
 with extensions varying from
$~0.1\deg$ to $~0.25\deg$ (\cite{hess_plane_survey}). In fact, within
our Galaxy only very few sources (most notably the $\gamma$-ray binaries
LS\,5039 (\cite{HESSLS5039}), LS I+61-303 (\cite{MAGICLSI, VERITASLSI})
HESS\,J0632+057 (\cite{HESS0632}), and the Crab nebula (\cite{crab_weekes})
are amongst the few Galactic sources
for which no significant extension was detected at \tev energies. Given
these findings, it can be expected that if the source population at \gev
energies has similar properties, that at least some of the Fermi-LAT
sources in our Galaxy should be spatially resolvable at \gev energies.

\todo[inline]{Ask Stefan/Markus if they think we should get rid of
this note about what point soruces in the plane.}

\todo[inline]{add note about how high energy emission may be hadron,
and the origin of cosmic rays}

Being able to spatially resolve the \gev emission from a source is
important for several reasons. Because of the large PSF and the large
candidate source density in our Galaxy, source confusion is a severe
problem for the identification of LAT sources.  Finding a coherent
source extension across different energy bands allows a firm association
of a LAT source to an otherwise confused counterpart.  Because of the
strongly varying PSF of the LAT, the spatial and spectral information
about a source do not simply decouple. A biased spatial model will bias
the spectral model of the source. In particular, performing a spectral
analysis on a spatially extended source modeled as point-like will
systematically shift the spectral index to be softer.  Furthermore,
correctly modeling a source's extension is important for improving
the model of the sky and removing excess residuals, for example in the
region around surrounding the Large Magellanic Cloud (\cite{first_cat}).
Additionally, such excess residuals potentially bias the significance
and measured spectrum of neighboring sources in the densely populated
Galactic plane.

@@

Morphological studies of sources in the \gev energy range using the
LAT are challenging because of the significantly energy-dependent PSF.
Additional complications arise close to the Galactic plane due to
systematic uncertainties in the Galactic diffuse emission.  The LAT's
PSF is limited at lower energies by multiple scattering in the tracker
and the 68\% containment radius of the PSF is approximately 5.1\deg
at 100 \mev  when averaged over instrument acceptance and including
photons which convert in either the thick or thin layers of the
tracker. The PSF improves with energy and approaches a limit given
by the granularity of the silicon strips and is 0.14\deg at 10 \gev
(\cite{on_orbit_calibration}).  The high resolution of the higher
energy photons in the LAT energy range however is offset by their low
statistics. Therefore a detailed investigation has been performed in this
work to determine the optimal energy range to measure source extension
(see below).

@@

This analysis is not feasible using the standard LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center (\cite{fssc})} because \gtlike can only 
fit the spectral parameters of the model.

We note that \gtlike has been used in the past in several studies of
source extension in the LAT collaboration (\cite{lmc,smc,w28,w51c}).
Based upon a set of \gtlike maximum likelihood fits at fixed extensions,
a profile of the likelihood as a function of the extension was calculated.
This approach is not optimal because the position, extension, and
spectrum of the source must be simultaneously fit to find the best fit
parameters and to correctly compute the statistical significance of the
detection.  Furthermore since the \gtlike likelihood profile approach
is computationally intensive, no large-scale Monte Carlo effort can be
run to validate it.

The approach presented here is based on a second maximum likelihood
fitting package developed in the LAT collaboration and called \pointlike
(\cite{first_cat,matthew_kerr_thesis}).  We extended the code to allow
a simultaneous fit of the source extension together with the position
and the spectral parameters.

The choice to base the spatial extension fitting on \pointlike
rather than \gtlike was made on considerations of computing time.
The \pointlike algorithm was optimized for speed to handle
larger numbers of sources efficiently which is important for
our catalog scan as well as the Monte Carlo validation of our method.
Details on the \pointlike package can be found in (\cite{matthew_kerr_thesis}).

@@

In \pointlike, it is assumed that the spatial and spectral
model of an extended source are separable, i.e. that the source
model $M(l,b,E)=S(l,b)\times X(E)$.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a
function of energy) and uses the \minuit library to maximize the
likelihood by simultaneously varying the position and extension of
the source (\cite{minuit_documentation}).  As will be described in
section~\ref{monte_carlo_validation}, simultaneously fitting the position
and extension is necessary to correctly calculate the statistical
significance of the detection of extension.  For each position and
extension, the spectral parameters of the sky model are refit.  To avoid
projection effects, the fitted source position parameters are not the
source's longitude and latitude but instead the source's displacement
in a rotated reference frame.

@@

For extended sources with an assumed radially symmetric shape,
we use a semi-analytic calculation to speed up the calculation of
the expected photon distribution. The expected photon 
distribution can be written as
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') r'^2 dr' d\phi'
\end{equation}
For the LAT, the PSF can be parameterized by a {\em King function}
\begin{equation}
  \text{PSF}(r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $u=(r/\sigma)^2/2$ and $\sigma$ and $\gamma$ are free parameters
(\cite{matthew_kerr_thesis}).  For radially symmetric extended sources,
the angular part of the integral can be done analytically
\begin{align}
  \text{PDF}(u) & = \int_0^\infty r'^2 dr'
  I_\text{src}(v) 
  \int_0^{2\pi} d\phi' 
  \text{PSF}(\sqrt{2\sigma^2(u+v-2\sqrt{uv}\cos(\phi-\phi'))})
  \\
  & = \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times {}_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{align}
where $v=(r'/\sigma)^2/2$ and ${}_2F_1$ is the Gaussian hypergeometric
function.  This convolution formula reduces the expected photon
distribution to a single numerical integral.

@@

We define \tsinc as twice the increase in \loglikelihood fitting the
region as two point sources compared to fitting the region as one point
source: 
\begin{equation}
  \tsinc=2\times(\loglikelihood_\text{2pts}-\loglikelihood_\text{ps}).
\end{equation} 
\tsinc can not be directly compared to \tsext to see which
model is more significant because the models are not nested
(\cite{statistics_with_care}). Even though the comparison of \tsext with
\tsinc is not a calibrated test, we find the cases $\tsinc \ll \tsext$
or $\tsinc\gg\tsext$ suggestive and we only consider a source to be
extended if $\tsext>\tsinc$.
Like extended sources described in section~\ref{gtlike_crosscheck},
the spectrum of the two point sources can be refit using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.

@@

Mattox et al. discuss that the test statistic distribution
for a likelihood ratio test on the existence of a source at
a given position is \cite{mattox_egret}
\begin{equation}\label{ts_ext_distribution}
  P(\ts)=\tfrac{1}{2}(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
The particular form of equation \ref{ts_ext_distribution} is
due to the null hypothesis (source flux $\Phi=0$) residing
on the edge of parameter space and the model hypothesis
adding a single degree of freedom. It is plausible to
expect a similar distribution of the test statistic
in the test for source extension as the same conditions
apply (with source flux $\Phi$ replaced by source radius $r$).
To validate this claim, we performed a Monte Carlo
study to calculate empirical distributions for $\tsext$ and
compared them to equation~\ref{ts_ext_distribution}.

We simulated point sources of varying spectral models using
\gtobssim\footnote{\gtobssim is distributed by the Fermi Science Support
Center (\cite{fssc}).} and fit them them using \pointlike as both point
and extended sources.  The point sources were simulated with a power-law
spectral model with integrated fluxes above 100 \mev ranging from $3\times
10^{-9}$ to $10^{-6} \ph/\cm^2/\sec$ in six discrete steps and spectral
indices ranging from 1.5 to 3 in four discrete steps.  These values
were picked to to represent typical source parameters in LAT-detected
sources. The point sources were simulated on top of an isotropic
background with an integrated flux above 100 \mev of $1.5\times10^{-5}$
(\cite{sreekumar_isotropic}).  The Monte Carlo simulation was performed
over a one year observation period using a default rocking profile and a
representative livetime fraction of 0.8.  The reconstruction was performed
using 1-100 \gev photons and the Pass 7\_V6 (P7\_V6) Source Instrument
Response Function (IRFs).  For each point source, we used \pointlike
to fit it as both a point and extended source and calculate \tsext and
we only kept the sources which had a significant point source detection
($\ts>25$).

@@

Figure~\ref{all_sensitivity}
shows the flux threshold as a function of source extension for
different background levels (1x, 10x, and 100x the
nominal background), different spectral indices, and two
different energy bands (1-100 \gev and 10-100 \gev).
The detection threshold is higher for sources in regions
of higher background.
For a fixed 1-100 \gev or 10-100 \gev
flux, the LAT's detection threshold is only weakly depends upon the
spectral index.  
Overlaid on figure~\ref{all_sensitivity} are the LAT
detected extended sources from table~\ref{known_extended_sources} and
table~\ref{new_ext_srcs_table}.  
This effect is most pronounced when using only photons
with energies above 10 \gev. The extension thresholds are tabulated in
table~\ref{all_sensitivity_table}.

Finally, figure~\ref{time_sensitivity} shows the LAT's projected detection
threshold to extension after 10 years against 10 times the isotropic
background. This background is representative of the background near
the Galactic plane.  For small extended sources, our detection threshold
improves better than $\sqrt{\text{t}}$ because at high energies, where we
are most sensitive to extension, the background levels are in the Poisson
instead of Gaussian regime.  For large extended sources, the relevant
background is larger and the improvement is closer to $\sqrt{\text{t}}$.

@@

Our analysis setup is closely related to the setup used for the
derivation of the catalog We used the same two years of data from August
4, 2008 to August 1, 2010 and we used the same Pass 7\_V6 (P7\_V6)
Source class event selection and Instrument Response Functions (IRFs,
\cite{lat_on_orbit_psf}).  We used the same models to describe the
background from Galactic diffuse, isotropic, and earth limb emission.
The galactic diffuse emission was scaled by a power-law and the isotropic
component's normalization was left free.

@@

We also performed a search for extended sources using only 10-100 \gev photons. 
Even though we are testing the same
sources, this approach is complimentary because the Galactic diffuse
emission is even less dominant above 10 \gev. Furthermore, source
confusion becomes less of a problem since most LAT sources are not
significant at energies above 10 \gev.  Using higher energies, we might
expect to detect harder sources in more complicated regions. This is
beneficial for regions near pulsars which cutoff before 10 \gev and was
successfully used in detecting HESS\,J1825-137 and MSH 15-52 with the LAT
(\cite{msh1552,fermi_hess_j1825}).

@@

Based on size vs. distance considerations,
we expect most extended sources to be inside our Galaxy.  Unfortunately, the \gev emission
in the Galactic plane is dominated by extremely structured 
diffuse emission from the interactions of cosmic rays with the
intergalactic medium.  Finding sources on top of this emission is difficult
(\cite{first_diffuse_paper}) as has been discussed in 1FGL and 2FGL
(\cite{first_cat,second_cat}).  Furthermore, the Galactic plane is crowded
and it is often difficult to correctly model nearby sources.  Because of this,
finding a source with $\tsext>16$ is not a sufficient
criteria for resolving a source. For each extended source,
we perform several analysis crosschecks.

@@

We estimate a systematic error on the extension of a source due to
uncertainty in our knowledge of the angular resolution of the LAT.  Before
launch, the LAT PSF was determined by a detector simulation which was
verified in accelerator tests (\cite{atwood_LAT_mission}).  A discrepancy
was found in flight data above a few \gev in the PSF compared to the
shape of bright AGN.  A publication on this issue is in preparation
(\cite{lat_on_orbit_psf}).  Subsequently, the PSF was fit empirically to
bright AGN and this empirical parameterization is the default PSF used
in the P7\_V6 IRFs.  To account for this uncertainty in our knowledge of
the PSF, we refit our extended source candidates using the Monte Carlo
representation of the PSF and consider the difference in extension found
using the two PSFs as a systematic error on the extension of a source.
The same approach was used in
(\cite{ic443}).  At high energies, we believe that our parameterization
of the PSF from bright AGN is substantially better than the Monte Carlo
representation of the PSF so this systematic errors is conservative.

@@

We estimate a second systematic error on the extension of a source
due to uncertainty in the model of the Galactic diffuse emission by
using an alternative diffuse model. The alternative model was based on
the GALPROP\footnote{GALPROP is a software package for calculating the
Galactic $\gamma$-ray emission based on a model of cosmic-ray propagation
in the Galaxy. See \url{http://galprop.stanford.edu/} for details and
references} model used in the LAT analysis of the isotropic diffuse
emission (\cite{isotropic_lat}).  
The intensities of various contributions to the galactic
diffuse emission are then fitted individually using a spatial
distribution of the intensities within the ROI 
as predicted by the model.
We distinguish contributions from CR interactions with
the molecular hydrogen, the atomic+ionized hydrogen,
residual gas traced by dust (\cite{isabelle_dark_gass}),
and the interstellar radiation field. We further
split the contributions from interactions with molecular
and atomic hydrogen to the
Galactic diffuse emission according
to the distance from the Galactic center in
which they are produced. Hence, we replace
the standard diffuse emission model by 18
individually fitted templates to describe individual
components of the diffuse emission.
A similar crosscheck was used in the
LAT Collaboration's analysis of RX J1713.7-3946 (\cite{rx_j1713_lat}).

@@

To validate our method, we test LAT sources associated with AGN for
extension.  \gev emission from AGN is believed to originate from the cores
of kpc-scale jets of distant galaxies.  Therefore AGN are not expected
to be spatially resolvable and provide a good calibration source to
demonstrate the efficacy of our tool.  We note that mpc-scale $\gamma$-ray
halos around AGNs have been hypothesized (\cite{pair_halo_paper}) which
could be resolved by the LAT However, no such halo has been discovered
in the LAT data so far.

>>>>

\bibliographystyle{apj}
\newcommand{\etal}{et al.}

@@

% for highlighting text
\usepackage{framed}
\colorlet{shadecolor}{yellow}

@@

% some units
\newcommand{\kev}{\text{Kev}\xspace}
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\tev}{\text{TeV}\xspace}
\newcommand{\sr}{\text{sr}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\renewcommand{\sec}{\text{s}\xspace}
\newcommand{\tsext}{{\ensuremath{\text{TS}_{\text{ext}}}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_{\text{inc}}}\xspace}
\newcommand{\loglikelihood}{\ensuremath{LL}\xspace}
\newcommand{\chandra}{\text{{\em Chandra}}\xspace}
\newcommand{\swiftxrt}{\text{{\em Swift}/XRT}\xspace}
\newcommand{\rosat}{\text{{\em ROSAT}}\xspace}
\newcommand{\suzaku}{\text{{\em Suzaku}}\xspace}
\newcommand{\asca}{\text{{\em ASCA}}\xspace}
\newcommand{\einstein}{\text{{\em Einstein}}\xspace}
\newcommand{\xmmnewton}{\text{{\em XMM-Newton}}\xspace}

@@

Using two years of all sky surveying data, the LAT Collaboration published
a catalog of 1873 \gev sources \citep{second_cat}. Many of these can be
positionally associated with a variety of source classes including Active
Galactic Nuclei (AGN), pulsars, galaxies, and supernova remnants (SNRs).
While many of the LAT sources can be spatially resolved when observed
at other frequencies, detection of an extension at \gev energies by
the LAT is complicated by the rather large point-spread function (PSF)
of the instrument. The angular resolution for events converting in the
front part of the detector, defined as the 68\% containment radius for
a point source, is approximately $3.5^{\circ}$ at 100 \mev, improving
to about 0.1$^{\circ}$ at 10 \gev.

However, a number of sources positionally coincident with LAT sources
exhibit extension at other wavelengths which are larger or comparable to
the LAT PSF.  In particular certain SNRs, pulsar wind nebulae (PWNe),
molecular clouds, normal galaxies, galaxy clusters or dark matter
satellites can therefore be expected to be observed as spatially extended
sources in the LAT.  The LAT collaboration has previously reported on the
discovered of five spatially extended supernova remnants, IC443, W28, W44,
W51C, and RX J1713.7-3946 \citep{ic443,w28,w44,w51c,rx_j1713_lat}. In
addition, three extended PWNe were detected: MSH 15-52, Vela~X,
and HESS\,J1825-137 \citep{msh1552,velax,fermi_hess_j1825}, as well
as two close-by galaxies, the Large Magellanic and Small Magalenic
Cloud \citep{lmc,smc}, and finally the radio galaxy Centarus A
\citep{cen_a_lat}.

For each of these sources, the extension was determined based on
dedicated analysis of the particular source or region of interest
instead of being the outcome of a systematic search of the LAT catalog.
For the systematic scan of all LAT catalog sources presented here, we
therefore expect to find additional spatially extended LAT sources, in
particular since many extended sources in the Galactic plane have been
detected at \tev energies using Air Cherenkov detectors and \tev and
\gev emission often originates from the same population of high-energy
particles.  A variety of spatially extended sources have been detected
in the Galactic plane at \tev energies. Most prominent was a survey of
the Galactic plane using the High Energy Stereoscopic System (H.E.S.S)
which presented 14 spatially extended sources
 with extensions varying from
$~0.1\deg$ to $~0.25\deg$ \citep{hess_plane_survey}.
In fact, within
our Galaxy only very few sources (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61-303 \citep{MAGICLSI, VERITASLSI}
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes}
are amongst the few Galactic sources
for which no significant extension was detected at \tev energies. Given
these findings, it can be expected that if the source population at \gev
energies has similar properties, that at least some of the Fermi-LAT
sources in our Galaxy should be spatially resolvable at \gev energies.

\begin{shaded} 
High energy $\gamma$-rays are produced by the decay of $\pi^0$s produced
by hadronic interactions with interstellar matter and by relativistic
electrons due to Inverse Compton (IC) scattering and Bremsstrahlung
radiation.  Many of these spatially extended \tev sources in the
Galactic plane are unidentified. Some of these sources could produce
hadronic cosmic rays up to $\approx 10^{15}$ eV and subsequently radiate
$\gamma$-rays through $\pi^0$ decay \citep{blandford_and_eichler_1987}.
Studying the spectra of these \tev sources at \gev energies may help to
determine the emission mechanisms of these high energy sources.
\end{shaded}

Being able to spatially resolve the \gev emission from a source is
important for several reasons. Because of the large PSF and the large
candidate source density in our Galaxy, source confusion is a severe
problem for the identification of LAT sources.  Finding a coherent
source extension across different energy bands allows a firm association
of a LAT source to an otherwise confused counterpart.  Because of the
strongly varying PSF of the LAT, the spatial and spectral information
about a source do not simply decouple. A biased spatial model will bias
the spectral model of the source. In particular, performing a spectral
analysis on a spatially extended source modeled as point-like will
systematically shift the spectral index to be softer.  Furthermore,
correctly modeling a source's extension is important for improving
the model of the sky and removing excess residuals, for example in the
region around surrounding the Large Magellanic Cloud \citep{first_cat}.
Additionally, such excess residuals potentially bias the significance
and measured spectrum of neighboring sources in the densely populated
Galactic plane.

@@

Morphological studies of sources in the \gev energy range using the
LAT are challenging because of the strongly energy-dependent PSF.
Additional complications arise for sources along the Galactic plane due to
systematic uncertainties in the Galactic diffuse emission.  The LAT's
PSF is limited at lower energies by multiple scattering in the tracker
and the 68\% containment radius of the PSF is approximately 5.1\deg
at 100 \mev  when averaged over instrument acceptance and including
photons which convert in either the thick or thin layers of the
tracker. The PSF improves with energy and approaches a limit given
by the granularity of the silicon strips and is 0.14\deg at 10 \gev
\citep{on_orbit_calibration}.  However, since most high energy
astrophysical sources have power-law spectra, the better resolution of the higher
energy photons is offset by their low
statistics. Therefore a detailed investigation has been performed in this
work to determine the optimal energy range to measure source extension
(see below).

@@

This analysis is not feasible using the standard LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center \citep{fssc}} because \gtlike can only 
fit the spectral parameters of the model.

We note that \gtlike has been used in the past in several studies of
source extension in the LAT collaboration \citep{lmc,smc,w28,w51c}.
Based upon a set of \gtlike maximum likelihood fits at fixed extensions,
a profile of the likelihood as a function of the extension was calculated.
This approach is not optimal because the position, extension, and
spectrum of the source must be simultaneously fit to find the best fit
parameters and to correctly compute the statistical significance of the
detection.  Furthermore since the \gtlike likelihood profile approach
is computationally intensive, no large-scale Monte Carlo effort can be
run to validate it.

The approach presented here is based on a second maximum likelihood
fitting package developed in the LAT collaboration and called \pointlike
\citep{first_cat,matthew_kerr_thesis}.  We extended the code to allow
a simultaneous fit of the source extension together with the position
and the spectral parameters.

The choice to base the spatial extension fitting on \pointlike
rather than \gtlike was made on considerations of computing time.
The \pointlike algorithm was optimized for speed to handle
larger numbers of sources efficiently which is important for
our catalog scan as well as the Monte Carlo validation of our method.
Details on the \pointlike package can be found in \citep{matthew_kerr_thesis}.

@@

In \pointlike, it is assumed that the spatial and spectral
model of an extended source are separable, i.e. that the source
model $M(l,b,E)=S(l,b)\times X(E)$.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a
function of energy) and uses the \minuit library to maximize the
likelihood by simultaneously varying the position and extension of
the source \citep{minuit_documentation}.  As will be described in
section~\ref{monte_carlo_validation}, simultaneously fitting the position
and extension is necessary to correctly calculate the statistical
significance of the detection of extension.  For each position and
extension, the spectral parameters of the sky model are refit.  To avoid
projection effects, the fitted source position parameters are not the
source's longitude and latitude but instead the source's displacement
in a rotated reference frame.

@@

For extended sources with an assumed radially symmetric shape,
we use a semi-analytic calculation to speed up the calculation of
the expected photon distribution. The expected photon 
distribution can be written as
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') r'^2 dr' d\phi'
\end{equation}
For the LAT, the PSF can be parameterized by a {\em King function}
\begin{equation}
  \text{PSF}(r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $u=(r/\sigma)^2/2$ and $\sigma$ and $\gamma$ are free parameters
\citep{matthew_kerr_thesis}.  For radially symmetric extended sources,
the angular part of the integral can be done analytically
\begin{align}
  \text{PDF}(u) & = \int_0^\infty r'^2 dr'
  I_\text{src}(v) 
  \int_0^{2\pi} d\phi' 
  \text{PSF}(\sqrt{2\sigma^2(u+v-2\sqrt{uv}\cos(\phi-\phi'))})
  \\
  & = \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times {}_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{align}
where $v=(r'/\sigma)^2/2$ and ${}_2F_1$ is the Gaussian hypergeometric
function.  This convolution formula reduces the expected photon
distribution to a single numerical integral.

@@

We define \tsinc as twice the increase in \loglikelihood fitting the
region as two point sources compared to fitting the region as one point
source: 
\begin{equation}
  \tsinc=2\times(\loglikelihood_\text{2pts}-\loglikelihood_\text{ps}).
\end{equation} 
\tsinc can not be directly compared to \tsext to see which
model is more significant because the models are not nested
\citep{statistics_with_care}. Even though the comparison of \tsext with
\tsinc is not a calibrated test, we find the cases $\tsinc \ll \tsext$
or $\tsinc\gg\tsext$ suggestive and we only consider a source to be
extended if $\tsext>\tsinc$.
Like extended sources described in section~\ref{gtlike_crosscheck},
the spectrum of the two point sources can be refit using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.

@@

\cite{mattox_egret} discuss that the test statistic distribution
for a likelihood ratio test on the existence of a source at
a given position is 
\begin{equation}\label{ts_ext_distribution}
  P(\ts)=\tfrac{1}{2}(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
The particular form of equation \ref{ts_ext_distribution} is
due to the null hypothesis (source flux $\Phi=0$) residing
on the edge of parameter space and the model hypothesis
adding a single degree of freedom. It is plausible to
expect a similar distribution of the test statistic
in the test for source extension as the same conditions
apply (with source flux $\Phi$ replaced by source radius $r$).
To validate this claim, we performed a Monte Carlo
study to calculate empirical distributions for $\tsext$ and
compared them to equation~\ref{ts_ext_distribution}.

We simulated point sources of varying spectral models using
\gtobssim\footnote{\gtobssim is distributed by the Fermi Science Support
Center \citep{fssc}.} and fit them them using \pointlike as both point
and extended sources.  The point sources were simulated with a power-law
spectral model with integrated fluxes above 100 \mev ranging from $3\times
10^{-9}$ to $10^{-6} \ph/\cm^2/\sec$ in six discrete steps and spectral
indices ranging from 1.5 to 3 in four discrete steps.  These values
were picked to to represent typical source parameters in LAT-detected
sources. The point sources were simulated on top of an isotropic
background with an integrated flux above 100 \mev of $1.5\times10^{-5}$
\citep{sreekumar_isotropic}.  The Monte Carlo simulation was performed
over a one year observation period using a default rocking profile and a
representative livetime fraction of 0.8.  The reconstruction was performed
using 1-100 \gev photons and the Pass 7\_V6 (P7\_V6) Source Instrument
Response Function (IRFs).  For each point source, we used \pointlike
to fit it as both a point and extended source and calculate \tsext and
we only kept the sources which had a significant point source detection
($\ts>25$).

@@

\begin{shaded}
Figure~\ref{all_sensitivity} shows the flux threshold as a function of
source extension for different background levels (1x, 10x, and 100x the
nominal background), different spectral indices, and two different energy
bands (1-100 \gev and 10-100 \gev).  The detection threshold is higher
for sources in regions of higher background.  For a fixed 1-100 \gev or
10-100 \gev flux, the LAT's detection threshold is only weakly depends
upon the spectral index.  Overlaid on figure~\ref{all_sensitivity} are
the LAT detected extended sources from table~\ref{known_extended_sources}
and table~\ref{new_ext_srcs_table}.  This effect is most pronounced when
using only photons with energies above 10 \gev. The extension thresholds
are tabulated in table~\ref{all_sensitivity_table}.

Finally, figure~\ref{time_sensitivity} shows the LAT's projected detection
threshold to extension after 10 years against 10 times the isotropic
background. This background is representative of the background near
the Galactic plane.  For small extended sources, our detection threshold
improves better than $\sqrt{\text{t}}$ because at high energies, where we
are most sensitive to extension, the background levels are in the Poisson
instead of Gaussian regime.  For large extended sources, the relevant
background is larger and the improvement is closer to $\sqrt{\text{t}}$.
\end{shaded}

@@

Our analysis setup is closely related to the setup used for the
derivation of the catalog We used the same two years of data from August
4, 2008 to August 1, 2010 and we used the same Pass 7\_V6 (P7\_V6)
Source class event selection and IRFs \citep{lat_on_orbit_psf}.  We used the same models to describe the
background from Galactic diffuse, isotropic, and earth limb emission.
The galactic diffuse emission was scaled by a power-law and the isotropic
component's normalization was left free.

@@

We also performed a search for extended sources using only 10-100 \gev photons. 
Even though we are testing the same
sources, this approach is complimentary because the Galactic diffuse
emission is even less dominant above 10 \gev. Furthermore, source
confusion becomes less of a problem since most LAT sources are not
significant at energies above 10 \gev.  Using higher energies, we might
expect to detect harder sources in more complicated regions. This is
beneficial for regions near pulsars which cutoff before 10 \gev and was
successfully used in detecting HESS\,J1825-137 and MSH 15-52 with the LAT
\citep{msh1552,fermi_hess_j1825}.

@@

Based on size vs. distance considerations,
we expect most extended sources to be inside our Galaxy.  Unfortunately, the \gev emission
in the Galactic plane is dominated by extremely structured 
diffuse emission from the interactions of cosmic rays with the
intergalactic medium.  Finding sources on top of this emission is difficult
\citep{first_diffuse_paper} as has been discussed in 1FGL and 2FGL
\citep{first_cat,second_cat}.  Furthermore, the Galactic plane is crowded
and it is often difficult to correctly model nearby sources.  Because of this,
finding a source with $\tsext>16$ is not a sufficient
criteria for resolving a source. For each extended source,
we perform several analysis crosschecks.

@@

We estimate a systematic error on the extension of a source due to
uncertainty in our knowledge of the angular resolution of the LAT.  Before
launch, the LAT PSF was determined by a detector simulation which was
verified in accelerator tests \citep{atwood_LAT_mission}.  A discrepancy
was found in flight data above a few \gev in the PSF compared to the
shape of bright AGN.  A publication on this issue is in preparation
\citep{lat_on_orbit_psf}.  Subsequently, the PSF was fit empirically to
bright AGN and this empirical parameterization is the default PSF used
in the P7\_V6 IRFs.  To account for this uncertainty in our knowledge of
the PSF, we refit our extended source candidates using the Monte Carlo
representation of the PSF and consider the difference in extension found
using the two PSFs as a systematic error on the extension of a source.
The same approach was used in
\citep{ic443}.  At high energies, we believe that our parameterization
of the PSF from bright AGN is substantially better than the Monte Carlo
representation of the PSF so this systematic errors is conservative.

@@

We estimate a second systematic error on the extension of a source
due to uncertainty in the model of the Galactic diffuse emission by
using an alternative diffuse model. The alternative model was based on
the GALPROP\footnote{GALPROP is a software package for calculating the
Galactic $\gamma$-ray emission based on a model of cosmic-ray propagation
in the Galaxy. See \url{http://galprop.stanford.edu/} for details and
references} model used in the LAT analysis of the isotropic diffuse
emission \citep{isotropic_lat}.  
The intensities of various contributions to the galactic
diffuse emission are then fitted individually using a spatial
distribution of the intensities within the ROI 
as predicted by the model.
We distinguish contributions from CR interactions with
the molecular hydrogen, the atomic+ionized hydrogen,
residual gas traced by dust \citep{isabelle_dark_gass},
and the interstellar radiation field. We further
split the contributions from interactions with molecular
and atomic hydrogen to the
Galactic diffuse emission according
to the distance from the Galactic center in
which they are produced. Hence, we replace
the standard diffuse emission model by 18
individually fitted templates to describe individual
components of the diffuse emission.
A similar crosscheck was used in the
LAT Collaboration's analysis of RX J1713.7-3946 \citep{rx_j1713_lat}.

@@

To validate our method, we test LAT sources associated with AGN for
extension.  \gev emission from AGN is believed to originate from the cores
of kpc-scale jets of distant galaxies.  Therefore AGN are not expected
to be spatially resolvable and provide a good calibration source to
demonstrate the efficacy of our tool.  We note that mpc-scale $\gamma$-ray
halos around AGNs have been hypothesized \citep{pair_halo_paper} which
could be resolved by the LAT However, no such halo has been discovered
in the LAT data so far.
